Kindly Bent to Free Us

### 2019-07-11 kindly/soa.tex
tighten the SOA a bit.

\subsection{Substructural type-system in functional languages}

@@

Quill~\citep{DBLP:conf/icfp/Morris16} is an Haskell-like language with affine
types through the use of qualified types.
Quill does not exposes a kind language, but
uses typeclass-like mechanism and annotation on arrows.
It supports both a most general type for function composition and
principal type inference.
We believe every Quill's kind signature can be expressed in \lang by leveraging
kind inequalities.
Unlike Quill,
we provide a linearity-aware semantics and prove its soundness.
Quill does not support borrows.

@@

Alms~\citep{DBLP:conf/popl/TovP11} is an ML-like language with rich, kind-based
affine types and ML modules.
Their system often relies on existential types to track the identity
of objects. For instance, they present an array creation
functions of type \lstinline/int -> 'a -> \E 'b. ('a, 'b) array/ where
\lstinline/'a/ is the elements' type and \lstinline/'b/
is a unique identifier of the array.
Due to the reliance on existentials, Alms does not support complete type inference.
Furthermore, Alms does not support borrow and often rely
on manual passing of capabilities.
In our experience, \affe's limited support for existential types through
regions is sufficient for a large majority of examples and provides
a more convenient programming style for imperative code.
%
Apart from existential types, Alms relies
on union and intersection operators and dependent kinds while
\lang uses qualified types.
We believe most of Alms' kind signatures can be expressed equivalently in
our system: for instance the pair type constructor
has kind $\Pi\alpha\Pi\beta. \langle\alpha\rangle \sqcup \langle\beta\rangle$
(where $\alpha$ and $\beta$ are types and $\Pi$ is the dependent function)
in Alms and $\kvar\to\kvar\to\kvar$ in \lang thanks
to subkinding.
%
Finally, Alms provides excellent support for abstraction through
modules by allowing to keep some type unrestricted inside a module, but
exposing it as affine. We support a similar programming style, which
we showcase in \cref{motivation}.

Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18} aims
to retrofit Haskell with linear types.
Linear Haskell introduces a new notion
of ``linear arrows'', noted $\multimap$ similar to linear logic,
which are functions who use their argument exactly once.
% By contrast, in \lang linearity is only decided
% by the kinds, and $\tarr{\klin}$ simply denotes \emph{single-use functions}.
This design is easy to retrofit on top of an existing compiler
such as GHC, but has several disadvantages\footnote{
  See the in-depth discussion attached to the GHC proposal starting here: \url{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}}.
Most relevant to \lang:
\begin{itemize}
\item Their system does not admit subtyping for arrows and leverages
  $\eta$-expansion to allow passing unrestricted functions in linear
  contexts. This is acceptable in a non-strict language such as
  Haskell but can not be used in a strict setting.
\item
  While the GHC proposal implements some type inference,
  their algorithm is not formalized and is not complete.
\item
  Linear Haskell promotes a CPS style with functions such as 
  \lstinline/withFile : path -> (file ->. file) ->. unit/
  to ensure that files are only used linearly.
  \lang follows System F\degree, Quill and Alms and
  adopts a direct style, where types themselves are
  described as affine or linear.
  We believe this leads to a more natural programming style
  and ease modular reasoning about linearity.
  In particular, using abstraction through modules,
  programmers only need consider the module
  implementation to ensure that some linear
  resource is properly handled.
% \item
%   Linear Haskell introduces a notion of linear monad to express
%   imperative code conveniently. Again, this solution is suitable in Haskell,
%   but less appropriate in other contexts. Our borrow system allow
%   to provide a more free-form programming style for imperative code.
% \item
%   Due to the previous remark, ensuring that a given object is never
%   aliased is fairly difficult in Linear Haskell, as one would need to ensure
%   that no non-linear function can manipulate it. On the other
\end{itemize}

@@

\subsection{Other substructural systems}

@@

Rust~\citep{rust} is the first
mainstream language to popularize the idea of borrowing and ownership
for safe low-level programming.
\lang heavily draws inspiration from Rust's borrowing system to apply
it in a functional setting with type inference, a GC, and
an ML module system.
Rust's lifetime system is more explicit and more expressive than \lang,
however Rust does not provide type inference
and only provides \emph{partial} lifetime inference.
Recently, \citet{DBLP:journals/corr/abs-1903-00982} presented Oxide,
a core language for Rust which formalizes its ownership discipline
including non-lexical lifetimes, which we look forward to use to improve
our borrowing system.

Vault~\citep{DBLP:conf/pldi/DeLineF01}
is a low-level language which uses typestate along with \emph{keys}
to track ownership. It also introduces a novel paradigm
called \emph{adoption and focus} \citep{DBLP:conf/pldi/FahndrichD02}
which allow restricting the scope of linear objects
(adoption) or temporarily treating them as non-linear (focus).
These mechanisms serve a similar purpose as borrows, but are tracked
much more precisely thanks to the use of typestate and pre/post-conditions,
which makes it very expressive (at the cost of some verbosity).
However, Vault is mostly geared towards traditional imperative programming:
functions can't capture linear objects and it does not support
type inference.

Plaid \&Co by Aldrich and others

>>>>

\subsection{Substructural type-systems in functional languages}

@@

Quill~\citep{DBLP:conf/icfp/Morris16} is an Haskell-like language with affine
types through the use of qualified types.
Quill does not exposes a kind language, but
uses typeclass-like mechanism and annotation on arrows.
It supports both a most general type for function composition and
principal type inference.
We were able to express the type signatures of all Quill examples
in \lang by leveraging kind inequalities.
Unlike Quill,
we provide a linearity-aware semantics and prove its soundness.
Quill does not support borrows.

@@

Alms~\citep{DBLP:conf/popl/TovP11} is an ML-like language with rich, kind-based
affine types and ML modules, similar to \lang.
Their system often relies on existential types to track the identity
of objects. For instance
\lstinline/Array.create : int -> 'a -> \E 'b. ('a, 'b) array/ where
\lstinline/'b/ uniquely identify of the array.
Due to the reliance on existentials, Alms does not support complete type inference.
Furthermore, Alms does not support borrow and often rely
on manual passing of capabilities.
In our experience, \affe's limited support for existential types through
regions is sufficient to express Alms' examples and leads to
a more convenient programming style for imperative code.
%
Kind-wise, Alms relies
on unions, intersections and dependent kinds while
\lang uses qualified types.
We believe most of Alms' kind signatures can be expressed equivalently in
our system: for instance the pair type constructor
has kind $\Pi\alpha\Pi\beta. \langle\alpha\rangle \sqcup \langle\beta\rangle$
(where $\alpha$ and $\beta$ are types and $\Pi$ is the dependent function)
in Alms and $\kvar\to\kvar\to\kvar$ in \lang thanks
to subkinding.
%
Finally, Alms provides excellent support for abstraction through
modules by allowing to keep some type unrestricted inside a module, but
exposing it as affine. We support a similar programming style, which
we showcase in \cref{motivation}.

Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18} aims
to retrofit Haskell with linear types.
Linear Haskell introduces a new notion
of ``linear arrows'', noted $\multimap$ similar to linear logic,
which are functions who use their argument exactly once.
% By contrast, in \lang linearity is only decided
% by the kinds, and $\tarr{\klin}$ simply denotes \emph{single-use functions}.
This design is easy to retrofit on top of an existing compiler
such as GHC, but has several disadvantages\footnote{
  See the in-depth discussion attached to the GHC proposal starting here: \url{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}}.
Most relevant to \lang:
\begin{itemize}[leftmargin=*]
\item Their system does not admit subtyping for arrows and leverages
  $\eta$-expansion to pass unrestricted functions in linear
  contexts. This works in a non-strict language such as
  Haskell but can not be used in a strict setting.
\item
  While the GHC proposal implements some type inference,
  it is neither formalized nor complete.
\item
  Linear Haskell promotes a continuation-passing style with functions such as 
  \lstinline/withFile : path -> (file ->. file) ->. unit/
  to ensure that files are only used linearly.
  \lang follows System F\degree, Quill and Alms and
  adopts a direct style, where types themselves are
  described as affine or linear.
  We believe this leads to a more natural programming style
  and ease modular reasoning about linearity.
  In particular, using abstraction through modules,
  programmers only need to consider the module
  implementation to ensure that linear
  resources are properly handled.
% \item
%   Linear Haskell introduces a notion of linear monad to express
%   imperative code conveniently. Again, this solution is suitable in Haskell,
%   but less appropriate in other contexts. Our borrow system allow
%   to provide a more free-form programming style for imperative code.
% \item
%   Due to the previous remark, ensuring that a given object is never
%   aliased is fairly difficult in Linear Haskell, as one would need to ensure
%   that no non-linear function can manipulate it. On the other
\end{itemize}

@@

\subsection{Other substructural type-systems}

@@

Rust~\citep{rust} is the first
mainstream language to popularize the idea of borrowing and ownership
for safe low-level programming.
\lang heavily draws inspiration from Rust's borrowing system to apply
it in a functional setting with type inference, a GC, and
an ML module system.
Rust's lifetime system is more explicit and more expressive than \lang,
however Rust does not provide type inference
and only provides \emph{partial} lifetime inference.
Recently, \citet{DBLP:journals/corr/abs-1903-00982}
formalized Rust's ownership discipline, including non-lexical lifetimes.
We believe this work
could help us improve our handling of regions and of successive exclusive borrows.

Vault~\citep{DBLP:conf/pldi/DeLineF01}
and Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09}
leverage typestate and capabilities
to express very rich properties in objects and protocols.
These systems are designed for either low-level or object-oriented
programming, and do not immediately lend themselves to a more functional
style. Furthermore, while their type-systems are much more
powerful than \affe's, they require many annotations by the programmer
and do not support any inference.
It could however be interesting to expand \affe with limited
forms of typestate as a local, opt-in feature that would provide
more expressivity at the cost of inference.
